---
title: "Multivariate Analysis"
author: "Shukry Zablah"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: 
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Imports

```{r}
library(dplyr)
library(mosaic)
library(ggplot2)
library(ggcorrplot)
```

## Load Data

```{r}
PIMA <- readRDS(file = "../data/PIMA_noNAs.Rds")
```

Note that the following analysis is done for the dataset without any missing values. 

## Correlation of Variables

```{r}
PIMA %>%
  select(-hasDiabetes) %>%
  cor() %>% 
  ggcorrplot()
```

In the correlation plot above we can see that there are some featured that are correlated. This is a hint that we might now need both features of a correlated pair in our model as they are likely to not add valuable information. We can see that the stronger correlated features are: 

- insulin and glucose concentration:  `r with(PIMA, cor(insulin, glucoseConcentration))`
- age and pregnancies:  `r with(PIMA, cor(age, pregnancies))`
- bmi and skin thickness:  `r with(PIMA, cor(bmi, skinThickness))`

## Multivariate Visualizations

### Insulin and Glucose Concentration

```{r}
# viz for insulin vs glucoseConcentration (maybe include hasDiabetes?)
```

### Age and Pregnancies

```{r}
# viz for age vs pregnancies (maybe include hasDiabetes?) 
```

### BMI and Skin Thickness

```{r}
# viz for bmi and skin thickness (maybe include hasDiabetes?)
```

