---
title: "Wrangle"
author: "Shukry Zablah"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Imports

```{r message=FALSE}
library(dplyr)
library(tidyr)
library(caTools)
```

## Load Data

```{r}
newPIMA <- readRDS(file = "../data/PIMA_original.Rds")
```

## Modify Data

### Create Variable Names

```{r}
names(newPIMA) <- c("pregnancies", "glucoseConcentration", "bloodPressure", "skinThickness", "insulin", "bmi", "diabetesPedigreeFunction", "age", "hasDiabetes")
```

### Make the Response a Factor

```{r}
newPIMA <- newPIMA %>% 
  mutate(hasDiabetes = as.factor(hasDiabetes))
```

### Mark Missing Values

In this dataset, there are missing values marked as 0. An example of how we can identify this by noting that a value of 0 for blood pressure does not make sense.

```{r}
newPIMA <- newPIMA %>%
  mutate(glucoseConcentration = ifelse(glucoseConcentration == 0, NA_integer_, glucoseConcentration)) %>%
  mutate(bloodPressure = ifelse(bloodPressure == 0, NA_integer_, bloodPressure)) %>%
  mutate(skinThickness = ifelse(skinThickness == 0, NA_integer_, skinThickness)) %>%
  mutate(insulin = ifelse(insulin == 0, NA_integer_, insulin)) %>%
  mutate(bmi = ifelse(bmi == 0, NA_integer_, bmi))
```

## Impuation of Missing Values

```{r}
PIMA_noNAs <- PIMA %>% 
  drop_na()

dim(PIMA_noNAs)
```
There are 392 observations that don't have missing values. That's half of the data. We have to find a way to replace the missing values by sensive alternatives. For now, all work is done assuming we ignore the missing values.

## Test/Train Split

```{r}
set.seed(100)

split <- with(PIMA_noNAs, 
              sample.split(hasDiabetes, SplitRatio = 0.75))

PIMA_train <- subset(PIMA_noNAs, split == TRUE)
PIMA_test <- subset(PIMA_noNAs, split == FALSE)
```

## Save Files

```{r}
saveRDS(newPIMA, file = "../data/PIMA_wrangled.Rds")
saveRDS(PIMA_noNAs, file = "../data/PIMA_noNAs.Rds")
saveRDS(PIMA_train, file = "../data/PIMA_train.Rds")
saveRDS(PIMA_test, file = "../data/PIMA_test.Rds")
```